---
toc: true
bibliography: references.bib
csl: american-medical-association-brackets.csl
---

# Clinical Core Methods

Last Revised: October 10, 2024  
VERSION 1.0

```{r code block for footnotes}
#| echo: false
gen_fn <- function(n, id) {
  # n is number of footnote
  # id is a unique id for that footnote
  paste0(
    '<a href="#',
    id,
    '" class="footnote-ref" role="doc-noteref" aria-expanded="false"><sup>',
    n,
    '</sup></a>'
  )
}
```

## Background

This document provides information about the ABC-DS study and documentation of
data currently available on the LONI website. The initial ABC-DS study began in
2015. Funded by the National Institutes on Health (NIH) through the **U01**
mechanism, from 2015-2020, ABC-DS was the combined/harmonized effort of two
groups of investigators: 1) the Alzheimer’s Dementia in Down Syndrome (**ADDS**)
group (Columbia University/NYS Institute for Basic Research in Developmental
Disabilities, Massachusetts General Hospital, University of California Irvine);
2) the Neurodegeneration in Aging Down Syndrome (**NiAD**) group (University of
Pittsburgh, University of Wisconsin Madison, University of Cambridge, Banner
Institute, Washington University).

From 2020-2026, the ABC-DS study was funded by NIH through the **U19** mechanism.
The number of participating institutions increased to ten (Columbia
University/NYS Institute for Basic Research in Developmental Disabilities,
Massachusetts General Hospital, University of California Irvine, University of
Pittsburgh, University of Wisconsin Madison, University of Cambridge, Washington
University, Kansas University Medical Center, and the University of Puerto
Rico). A single, harmonized protocol, including many of the ADDS and NiAD
measures, was adopted by all ten institutions. The U19 project includes new
participants as well as individuals continuing from the U01 projects.

While the focus of ABC-DS is on individuals with Down Syndrome, NiAD also
recruited sibling controls, and unaffected siblings continue to be recruited in
the U19. In the ongoing U19 study, in addition to new participants, as many
former U01 participants as possible have been recruited, for continued
evaluation. 

ABC-DS participants are scheduled to be seen every 16 months. Sibling controls
receive a shorter cognitive evaluation battery than participants with Down
Syndrome and do not receive caregiver assessments.

## Available Data 

ABC-DS data has been harmonized and made available to the research community
through the Laboratory of Neuroimaging website (LONI). Cumulative data released
at annual intervals include evaluations in which a consensus diagnosis has been
generated as of a set date, based on a defined battery of tests. Currently, data
on a total of 1,274 evaluations conducted on 503 participants with Down Syndrome
(cases) are available on LONI; on 503 individuals with Down Syndrome and 59
sibling controls, along with 161 evaluations conducted on 59 sibling controls.
Case and control data are provided in separate tables.

Data collected at multiple evaluations are organized in a
one-record-per-evaluation format sorted by participant ID, subject label, and
consecutive evaluation (event_sequence), regardless of study of recruitment. For
example, a participant with three evaluations in the U01 and one evaluation in
the U19 would have data for event_sequence 1, 2, 3 and 4 (the first U19
evaluation). A participant seen in the first and third U01 evaluation with one
evaluation in the U19 would have data for event_sequence 1,2 and 3. Note that
event_sequence is numbered consecutively, regardless of whether a particular 16
month month evaluation (in this case the second U01 evaluation) is missed. The
age_at_event_and_latency table provides the age of the participant at each
evaluation and the latency in days from baseline of each subsequent evaluation.
The number of evaluations for which data are currently available is shown in
Table 1. Please note that the N of evaluations available for any specific
measure is affected by a number of factors (please see descriptions of
individual measures, below).


*Table 1. Number of available ABC-DS evaluations*

```{r}
#| echo: false
#| ft.align: left
columns <- c(
  "Evaluation (event_sequence)",
  "Down Syndrome Cases",
  "Sibling Controls"
)
evaluation <- c(
  "1st (baseline)",
  "2nd evaluation",
  "3rd evaluation",
  "4th evaluation",
  "5th evaluation",
  "Total evaluations"
)
ds <- c(503, 346, 283, 138, 4, 1274)
controls <- c(59, 35, 17, 5, 1, 116)

data.frame(evaluation, ds, controls) |>
  setNames(columns) |>
  flextable::flextable() |>
  flextable::autofit() |>
  flextable::align(align = "center", part = "all")
```

## ABC-DS Cognitive & Caregiver Measures (2015-2026)

Some measures in the harmonized data available on LONI were collected in the U19
study, but not in one or both of the U01studies (ADDS and/or NiAD); for the
tables containing these measures (flagged in the descriptions below), sample
sizes will be smaller than for measures administered in all three studies.

## Measures in the ABC-DS Integrated Dataset

### Cognitive Assessment Measures

For all the cognitive measures, before performing any modelling, please check
the distributions of the variables. The vast majority of them have a skewed
distributions and are zero inflated, which requires either some type of
transformation or proper modelling.

a) The **Down Syndrome Mental Status Examination (DSMSE)** (an enhanced
version)[@haxby1989neuropsychological] provides an omnibus measure of
neuropsychological functioning, assessing recall of personal information,
orientation, immediate and delayed memory, language, visuospatial function and
praxis.

b) The Rapid Assessment for Developmental Disabilities (RADD)[@mccarthy1972msca]
Assesses the ability of the individual to copy simple hand movements. For ABC-DS
we are only using the Hand Movements and Imitation Subtest of the measure.`r gen_fn(2, "fn2")`

c) The McCarthy Scales of Children’s Abilities: Category Fluency
Test[@korkman2007nepsy] An adaptation of the McCarthy Category Fluency Test has
been employed to assess changes in language ability and frontal lobe function
over time. It requires participants to name as many foods and animals as they
can within 20 seconds. We have been using a single summary score generated by
adding the number of correct responses for each category.`r gen_fn(2, "fn2")`

d) The NEPSY (2nd ed.) Word Generation Semantic Fluency
Subtest[@korkman2007nepsy] assesses verbal productivity through the ability to
generate as many words as possible in 20 sec. within a specific semantic
category (e.g., animals, food and drinks). The score is the total number of
unique animal names/foods/drinks generated.`r gen_fn(2, "fn2")`

e) Kaufman Brief Intelligence Test (2nd ed)[@kaufman2004kaufman] The KBIT-2 is
a brief measure of verbal and nonverbal intelligence for ages 4-90. It provides
crystallized (verbal) and Fluid (nonverbal) score along with an IQ composite
score.`r gen_fn(1, "fn1")`

f) The Wechsler Intelligence Scale for Children, 4th Edition
(WISC-IV)[@wechsler2003wisc] Block Design subtest was used to measure
visuo-spatial organization. The subtest is a core perceptual reasoning timed
test that requires that the individual reproduce designs using blocks that have
red surfaces, white surfaces, and surfaces divided diagonally into half red and
half white. The Digit span forward subtest of the WISC-IV was used to assess
rote learning and memory. The WISC-IV version is somewhat easier (as it is
designed for children ages 6 to 16). It contains series ranging in length from
two to nine digits and primarily involves rote learning and memory. Digit
forward requires participants to recite a series of digits in the exact order
given by the evaluator (forward). Conversely, Digit Span backwards requires
participants to recite series of digits backwards and taps ability to manipulate
verbal material in working memory.

g) Haxby Extension.[@haxby1989neuropsychological] Like the WISC Block Design
subtest, the Haxby Extension requires that participants reproduce designs using
blocks with red surfaces, white surface, and surfaced divided diagonally into
half red and half white. However, the designs are simpler than those on the
WISC, such as starting with reproducing a single block. Possible cores range
from 0-16.

h) The Beery-Buktenica Developmental Test of Visual Motor Integration
(VMI)[@beery2004beery] involves the copying of figures of progressive complexity
using paper and pencil. It is designed to assess the extent to which the
individual can integrate his/her visual and motor abilities. Scores range from 0
to 30 (higher scores indicating more correctly copied items).

i) The Modified Cued Recall Test (mCRT)[@devenny2002cued] is a list-learning
task that measure verbal learning and memory. The mCRT consists of a training
period in which 12 items are presented, 4 at a time, with each item accompanied
by a unique category cue. The testing phase consists of 3 trials which generates
a free recall score (FRS -spontaneous recall of the 12 items), cued recall score
(CRS-items only recalled when the category cue is provided), a total score (sum
FRS and CRS), and an intrusion score (items named that were not on the original
list).

j) The Modified Cats and Dogs Task[@ball2008executive] is an adapted version of
the Stroop Task and it assesses cognitive inhibition. Specifically, the task
measures an individual’s ability to inhibit a natural response (naming pictures
of a dog as “dog”) and to replace it with an incongruent one (naming pictures of
a dog as “cat”).

k) The Purdue Pegboard Test[@tiffin1948purdue] measures manual dexterity (left
and right hand separately) as a participant places pins/pegs in a pegboard with
2 parallel rows.

l) The Cancellation Task-Revised[@krinsky2008selective] measures processing
speed, visual selective attention, vigilance, perceptual speed and visuomotor
ability by asking participants to circle a target stimulus/picture on a page
containing both target pictures and distractor pictures.

m) The Tinetti Assessment Tool: Gait subtest[@tinetti1986performance] quantifies
gait in 7 domains by observing participants while walking. Those domains are:
Initiation of Gait, Step Length, Step Height/ Foot Clearance, Step Symmetry,
Step Continuity, Path, Trunk and Walking Stance.

Caregiver Interview Measures

a) The Dementia Questionnaire for People with Learning Disabilities
(DLD).[@eurlings2006dementia] The DLD consists of 50 items in 8 subscales,
including short-term memory, long-term memory spatial and temporal orientation,
speech, practical skills, mood, activity and interest, and behavioral
disturbances. The DLD generates two scores, one reflecting cognitive abilities
(DLD/SCS) and one reflecting social skills (DLD/SOS).

b) The NTG Early Detection Screen for Dementia (NTG-EDSD).[@esralew2017national]
The NTG-EDSD includes sections on behavioral/cognitive domains that could
reflect indications of dementia clinical progression. In the current ABC-DS
cohort, the DLD and NTG significantly differentiated participants who were
Unaffected from those with MCI-DS and Dementia.

c) The Vineland Adaptive Behavior Scales, 3rd edition
(VABS-3),[@sparrow2016vineland] Domain Level Parent/ Caregiver Form, provides a
standardized assessment of functional abilities for use with individuals (ages 0
– 90+) having intellectual and developmental disabilities. The Domain Level
Parent/Caregiver Form is administered as a semi-structured interview with the
parent or caregiver and focuses on home and family- life behavior. Adaptive
functioning within three domains, Communication, Daily Living Skills, and
Socialization, are assessed.`r gen_fn(1, "fn1")` `r gen_fn(2, "fn2")`

d) The Reiss Screen for Maladaptive Behavior[@reiss1994handbook] was created
specifically for use with individuals with ID to screen for possible depression,
psychosis and behavior problems. Caregivers are asked to rate 38 symptoms of
psychiatric behavior as being “No Problem”, “Problem”, or “Major Problem” for
the individual in question. Behaviors are categorized into 8 sub-domains
reflecting (a) aggressive behavior, autistic-like behavior, (c) psychosis, (d)
paranoia, (e)/(f) behavioral and physical symptoms of depression, (g) dependent
personality disorder, and (h) avoidant disorder.

e) The Neuropsychiatric Inventory (NPI)[@cummings1994neuropsychiatric] is a
validated instrument used to assess neuropsychiatric symptoms in dementia. The
questionnaire asks caregivers to retrospectively rate symptoms in 12
neuropsychiatric domains: delusions, hallucinations, agitation/ aggression,
dysphoria/depression, anxiety, euphoria/elation, apathy/indifference,
disinhibition, irritability/lability, aberrant motor behaviors, nighttime
behavioral disturbances, and appetite/ eating disturbance.`r gen_fn(2, "fn2")`

Health Status and Life Events

a) Demographics Questionnaire; includes year of participant’s birth; sex;
ethnicity and race; primary language; age of mother and of father at
participant’s birth; handedness; primary residence; level of independence;
activities (e.g., vocational training program, working in the community,
postsecondary education program).

b) Medical Health History Questionnaire; includes Psychiatric conditions (e.g.,
depressive disorder, anxiety disorder); Neurological conditions (e.g., dementia,
stroke, seizure disorder); Eyes/ears/nose/throat (e.g., cataracts, visual
impairment, hearing impairment); Cardiovascular conditions (e.g., congenital
heart disease, pacemaker); Respiratory conditions (e.g., asthma, pneumonia,
sleep apnea); Allergies; Dermatologic/Connective tissue conditions (e.g.,
alopecia, rosacea); Musculoskeletal conditions (e.g., osteoporosis; gout);
Endocrine/metabolic conditions (e.g., diabetes, Vitamin B12 deficiency);
Menstrual information (e.g., menopausal); Hormone replacement information;
Gastrointestinal conditions (e.g., celiac disease, irritable bowel syndrome);
Hematopoetic conditions (e.g., anemia); Cancer; Surgery; Covid-19; and Other
conditions;

c) Stress Index: Caregivers were asked about the presence or absence of
significant life events and were asked to rate how the participant handled the
experience.`r gen_fn(2, "fn2")`

## Special Cautions When Using this Dataset

**Modified Cued Recall Test**. Two sets of stimuli images (Set A/ Version 1 or
Set B/ Version 2) were used. The set used varied across site and data collection
cycle. Analyses conducted thus far suggest that set/version administered
influences participant performance. The following variables are unique to the
ABC-DS U19 study: `crset`, `ift1r`, `ift2r`, `ift3r`, `tifrreps`, `agfrin1`, `agfrin2`,
`agfrin3`, `tafi`, `c1t1`, `c1t2`, `c1t3`, `c2t1`, `c2t2`, `c2t3`, `c3t1`, `c3t2`, `c3t3`, `crc1`, `crc2`
and `crc3`.

**Down Syndrome Mental Status Examination (DSMSE)**: A component of the DSMSE, the
Shoebox Memory Test, was unique to the ADDS U01 and ABC-DS U19 studies; thus,
sample sizes for the `colshbx` and `dsto2` will be smaller than those for other
DSMSE variables. Similarly, variables `colshbxtime`, `dsme1time`, `colmemscr` and
`nonmemory` are unique to the ABC-DS U19 study; thus, sample sizes for these
variables will be smaller than those for other variables.

**Modified Cats and Dogs Test**: The variables, `nttrain` and `strain`, are new
and unique to the ABC-DS U19 study.

**NTG-Early Detection Screen for Dementia (NTG-EDSD)**: The Caregiver Concern
Questions, variables `ntspcon` and `ntspconage`, are unique to the ADDS U01 and
ABC-DS U19 studies.

**Purdue Pegboard Test**: Variables `pp_dh_tr` and `pp_ndh_tr` are unique to the
ABC-DS U19 study. The variable `de_handedness` can be used to determine
participants' dominant hand (1 = Right-handed, 2 = Left-handed, 3 =
Ambidextrous).

**Stress Index**: The variable `si_victim` is a new variable unique to the ABC-DS U19 study.

**Tinetti Gait Assessment**: The following variables are unique to the ABC-DS
U19 study: `tbagait`, `tbaslh`, `tbafc`, `tbass`, `tbasc`, `tbap`, `tbat`, and
`tbawt`.

**Verbal Fluency- McCarthy Category Fluency Test**. The “Exceptional Scoring
Method” is used. Variables `vffdexepr` and `vfanexepr` are unique to the ADDS U01
and ABC-DS U19 studies; the variable `vftotexepr` is unique to the ABC-DS U19
study.

```{=html}
<footer>
  <h2>Footnotes</h2>
  <ol  style="padding-left: 1rem;">
    <li id="fn1"><p>Baseline Only</p></li>
    <li id="fn2"><p>Data of this type not collected in all three studies; N may be lower than for other measures.</p></li>
  </ol>
</footer>
```

## Protocol for Sibling Controls

Approximately 50 sibling controls were recruited and underwent the same
procedures with the exception of no caregiver assessments and a small cognitive
screening battery. In addition, during the 2015-2020 U01 grant, only baseline
blood draws were conducted. Sibling controls provided medical histories and were
screened using the AD8[@galvin2005ad8] (a brief informant interview to detect
dementia) and the Montreal Cognitive Assessment (MoCA).[@nasreddine2005montreal]
The two screens were used to simply ensure that the siblings did not have
MCI/Dementia and are not included in the ABC-DS database.